---
title: "Plot examples from the DCWS launch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot examples from the DCWS launch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggrepel}
  %\VignetteDepends{camiller}
---

```{r setup, message=FALSE, warning=TRUE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE,
  dpi = 150,
  out.width = "100%"
)
```

As a showcase of many of the functions in `stylehaven`, this is a simplified redo of several charts from the 2021 DCWS launch event, using data that's bundled up with this package.

```{r libraries, message=FALSE, warning=FALSE}
library(stylehaven)
library(ggplot2)
library(dplyr)
library(showtext)

showtext_auto()
showtext_opts(dpi = 150)
```

## Unified styling setup

Illustrates:

* Creating multiple weights of a font to use across all charts with `font_add_weights`
* Creating an array of color palettes based on a primary color

```{r style, fig.width=5, fig.height=3, message=TRUE, warning=TRUE}
font_add_weights("Barlow Semi Condensed", semibold = 500)

base_col <- "#5e3da7"
palx_colors <- palx(base_col, n_shades = 5, plot = TRUE)
qual_pal <- palx_colors$shade03
# use with offset labels to make nearly-white labels inside bars, dark gray labels outside
tf_pal <- c("TRUE" = palx_colors[["shade02"]][["gray"]], "FALSE" = palx_colors[["shade05"]][["gray"]])

theme_bar <- function(x, ...) {
  camiller::theme_din(base_family = "Barlow Semi Condensed", ygrid = FALSE) +
    theme(plot.caption.position = "plot", 
          strip.text = element_text(family = "Barlow Semi Condensed Semibold", face = "plain"))
}
theme_set(theme_bar())
update_geom_defaults("col", list(fill = base_col))
update_geom_defaults("text", list(size = 3.5, fontface = "bold", family = "Barlow Semi Condensed", color = tf_pal[["FALSE"]]))

pal4 <- unname(qual_pal[c("gray", "blue", "teal", "red")])
```

I'll keep all the plots in one named list, and all their output parameters (width & height, could also include something like logo placement) in another named list. Then I'll map over these to save all the plots as PDF and PNG. Having a list of widths & heights also makes it easier to have the same dimensions within this document (as chunk `fig.width` and `fig.height`) as they have once they're exported. Doing this prints an error when running chunks for some reason but you can ignore it.

```{r plot-setup}
plots <- list()
plot_params <- list(
  financial_insecurity = list(w = 6, h = 5),
  govt_trend = list(w = 5, h = 5),
  cohesion = list(w = 5, h = 4)
)
```


## Facetted bar chart

Illustrates:

* Offsetting labels across multiple scales with `offset_lbls`
* Lil bit of string formatting functions
* This won't show up when rendered, but I've set `fig.width=plot_params$financial_insecurity$w, fig.height=plot_params$financial_insecurity$h` for this chunk

```{r financial, fig.width=plot_params$financial_insecurity$w, fig.height=plot_params$financial_insecurity$h}
financial_to_plot <- fin_insecurity %>%
  filter(category %in% c("Connecticut", "Greater New Haven", "Age", "Race/Ethnicity")) %>%
  mutate(category = forcats::fct_collapse(category, Total = c("Connecticut", "Greater New Haven"))) %>%
  group_by(question) %>%
  offset_lbls(value = value, frac = 0.02, thresh = 0.2, fun = percent100)
financial_to_plot

plots[["financial_insecurity"]] <- ggplot(financial_to_plot, aes(x = forcats::fct_rev(group), y = value)) +
  geom_col(width = 0.8) +
  geom_text(aes(label = lbl, y = y, hjust = just, color = is_small), show.legend = FALSE) +
  coord_flip() +
  scale_y_barcontinuous(breaks = NULL) +
  scale_color_manual(values = tf_pal) +
  facet_grid(rows = vars(category), cols = vars(question),
             scales = "free", space = "free_y",
             labeller = labeller(.cols = camiller::clean_titles)) +
  theme(strip.text.x = element_text(hjust = 0),
        strip.text.y = element_blank()) +
  labs(x = NULL, y = NULL,
       title = "Younger adults and Latinos face hurdles to accessing basic needs",
       subtitle = "Share of adults, 2021")

plots[["financial_insecurity"]]
```

## Labeling endpoints

`endpoint_lbls` adds columns for labels that are offset slightly to the left for the lower limit and to the right for the upper limit, a corresponding hjust value to align left-hand labels at the end of the string and right-hand labels at the start of the string, and labels to use that include the grouping variable on one or both sides. This encourages direct labeling to wean yourself away from legends. 

One thing to deal with here is the margins of the plot: You can set padding with `expansion`, but that changes the limits of the scale as well, so you might need to adjust your breaks. That's easy enough for slopegraphs like this but could get annoying if you don't know exactly what the breaks will need to be.


```{r govt1, fig.width=plot_params$govt_trend$w, fig.height=plot_params$govt_trend$h}
govt_trend <- cws_trend %>%
  filter(indicator == "local_govt_responsive",
         year %in% range(year),
         category %in% c("Total", "Race/Ethnicity"))

(govt_left <- govt_trend %>%
  endpoint_lbls(x = year, value = value, group = group, 
                fun = percent100, long_side = "left"))

ggplot(govt_left, aes(x = year, y = value, color = group, group = group)) +
  geom_path(linewidth = 3) +
  geom_point(size = 4) +
  geom_text(aes(label = lbl, x = x, hjust = just)) +
  scale_x_continuous(expand = expansion(add = c(2, 1)), breaks = c(2015, 2021)) +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = pal4) +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "gray90")) +
  labs(x = NULL, y = NULL,
       title = "Views of local government are improving",
       subtitle = "Share of adults who agree that local government is responsive,\nConnecticut, 2015–2021")
```

If I want to switch sides for the long label, the only thing I need to change is the padding on the x-axis—everything else is in the data frame.

```{r govt2, fig.width=plot_params$govt_trend$w, fig.height=plot_params$govt_trend$h}
govt_right <- govt_trend %>%
  endpoint_lbls(x = year, value = value, group = group, 
                fun = percent100, long_side = "right") # this is the default side

plots[["govt_trend"]] <- ggplot(govt_right, aes(x = year, y = value, color = group, group = group)) +
  geom_path(linewidth = 3) +
  geom_point(size = 4) +
  geom_text(aes(label = lbl, x = x, hjust = just)) +
  # in practice I might use ggrepel
  scale_x_continuous(expand = expansion(add = c(1, 2)), breaks = c(2015, 2021)) +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = pal4) +
  theme(legend.position = "none",
        panel.grid.major.x = element_line(color = "gray90")) +
  labs(x = NULL, y = NULL,
       title = "Views of local government are improving",
       subtitle = "Share of adults who agree that local government is responsive,\nConnecticut, 2015–2021")

plots[["govt_trend"]]
```

For just a few lines, full labels on just the left or right is fine, but there's also the option to add labels on both sides, which can help make more lines easier to read

## Separating squished labels 

Dot plots with points directly labeled are great until the values are too close together, and then it's a pain to keep the labels from overlapping. One nice way to handle this is to calculate the distance between points on each line (whatever's the independent variable) and offset labels that are within some threshold of any other label. The data frame this returns can be used with `semi_join` and `anti_join` to make 2 calls to `geom_text`/`geom_text_repel`. The argument `thresh = 0.03` means values with a difference of 0.03 will be considered too close; this is an absolute value.


```{r dots1, fig.width=plot_params$cohesion$w, fig.height=plot_params$cohesion$h}
cohesion <- community_cohesion %>%
  filter(category %in% c("Greater New Haven", "Education"))

ggplot(cohesion, aes(x = question, y = value, color = group)) +
  geom_point(size = 9) +
  geom_text(aes(label = percent100(value)), size = 3, color = "white") +
  coord_flip() +
  scale_x_discrete(labels = camiller::clean_titles) +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = pal4),
                     guide = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1, linetype = "11", color = "gray90")) +
  labs(x = NULL, y = NULL, color = NULL)
```

```{r dots2}
(to_dodge <- cohesion %>%
  dodge_lbls(x = question, value = value, group = group, thresh = 0.02))
```

```{r dots3, fig.width=plot_params$cohesion$w, fig.height=plot_params$cohesion$h, eval=require("ggrepel")}
plots[["cohesion"]] <- ggplot(cohesion, aes(x = question, y = value, color = group)) +
  geom_point(size = 9) +
  geom_text(aes(label = percent100(value)), 
            data = . %>% anti_join(to_dodge, by = c("question", "group")),
            size = 3, color = "white") +
  ggrepel::geom_text_repel(aes(label = percent100(value)),
            data = . %>% semi_join(to_dodge, by = c("question", "group")),
            size = 3, fontface = "bold", family = "Barlow Semi Condensed",
            direction = "x", nudge_x = 0.2, min.segment.length = 2, seed = 1,
            segment.color = tf_pal[["TRUE"]], box.padding = 0.5) +
  coord_flip() +
  scale_x_discrete(labels = camiller::clean_titles) +
  scale_y_continuous(breaks = NULL) +
  scale_color_manual(values = pal4),
                     guide = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(size = 1, linetype = "11", color = "gray90")) +
  labs(x = NULL, y = NULL, color = NULL)

plots[["cohesion"]]
```

## Batch writing image files

After exporting about 5,000 plots for the equity reports, I'm excited to codify a way to go from the above lists of plots and plot parameters and not have to think about how to write all those plots out, adding & scaling logos. I didn't have a good way to test out lots of different graphics devices on different platforms, so file a Github issue if any exports don't work. 

By default, this will add a logo (using `cwi::add_logo`) and adjust the height of the plot to fit so that the aspect ratio stays the same even after appending the logo to the bottom of the image.

Writing out one plot (by default, both PNG & PDF):

```{r export1, message=TRUE, warning=FALSE}
out_dir <- file.path(tempdir(), "plots")
if (!dir.exists(out_dir)) dir.create(out_dir)

write_plot(plots$govt_trend, filename = "government_trend", 
           dir = out_dir, width = 5, height = 5, dpi = 150, verbose = FALSE)

list.files(out_dir, recursive = TRUE)
```






Even better, iterating over the plots and plot params to write all the plots. One thing to note here is to be careful that you're matching plots and params if they're in separate lists: you might get messed up if these lists are out of order, even if they're named. (This once bothered me enough to post a [question on Stack Overflow](https://stackoverflow.com/q/55798573/5325862).) You can map over one list and use its names to pull entries out of the other list.

```{r export2, message=TRUE, warning=FALSE}
purrr::iwalk(plots, function(pl, id) {
  ps <- plot_params[[id]]
  write_plot(pl, filename = id, dir = out_dir,
             width = ps$w, height = ps$h, dpi = 150, verbose = FALSE)
})

list.files(out_dir, recursive = TRUE)
```

